\documentclass{article}

% Pretty much all of the ams maths packages
\usepackage{amsmath,amsthm,amssymb,amsfonts}

% Set up some environments
\newtheorem{thm}{Theorem}[section]
\newtheorem{lem}[thm]{Lemma}
\newtheorem{dfn}{Definition}[section]
\theoremstyle{dfn}
\newtheorem{alg}{Algorithm}[section]
\numberwithin{equation}{section}
\numberwithin{figure}{section}

% Pretty much all of the algorithms packages
\usepackage{algorithmicx,algpseudocode}

% Allows you to manipulate the page a bit
\usepackage[margin=1in]{geometry}

% Pulls the page out a bit - makes it look better (in my opinion)
\usepackage{a4wide}

% Allows inclusion of graphics easily and configurably
\usepackage{graphicx}

% Provides ways to make nice looking tables
\usepackage{booktabs}

% Allows you to rotate tables and figures
\usepackage{rotating}

% Allows shading of table cells
\usepackage{colortbl}
% Define a simple command to use at the start of a table row to make it have a shaded background
\newcommand{\gray}{\rowcolor[gray]{.9}}

\usepackage{textcomp}

% Typesets URLs sensibly - with tt font, clickable in PDFs, and not breaking across lines
\usepackage{url}

% Makes references hyperlinks in PDF output
\usepackage{hyperref}

% Provides ways to include syntax-highlighted source code
\usepackage{listings}
\lstset{frame=single, basicstyle=\ttfamily}

% Provides good access to colours
\usepackage{color}
\usepackage{xcolor}

% Simple command I defined to allow me to mark TODO items in red
\newcommand{\todo}[1] {\textbf{\textcolor{red}{#1}}}

% Command for vector, matrix, tensor names
\renewcommand{\vec}[1] {\mathbf{#1}}
%\usepackage{esvect}
\newcommand{\Mat}[1] {\mathbf{#1}}
\newcommand{\Ten}[1] {\mathbf{\mathcal{#1}}}
\newcommand{\Pop}[1] {\mathcal{#1}}

% Command for important sets
\newcommand{\F} {\mathbb{F}}
\newcommand{\N} {\mathbb{N}}

% Math macros (NS)
\newcommand{\E} {\mathbb{E}}
\renewcommand{\epsilon}{\varepsilon}
\newcommand{\eps} {\epsilon}
% Allows fancy stuff in the page header
\usepackage{fancyhdr}
\pagestyle{fancy}

% Vastly improves the standard formatting of captions
\usepackage[margin=10pt,font=small,labelfont=bf, labelsep=endash]{caption}
\usepackage[margin=10pt,font=small,labelfont=bf, labelsep=endash]{subcaption}

\usepackage{cleveref}

% Standard title, author etc.
\title{An Efficient Fill Estimation Algorithm for Sparse Tensors in Blocked Formats}
\author{Peter Ahrens \and Nicholas Schiefer \and Helen Xu}
\date{}
% Put text on the left-hand and right-hand side of the header
%\fancyhead{}
%\lhead{}
%\rhead{}
%\chead{}

\begin{document}

  \maketitle

\begin{abstract}
Tensors, which are the linear-algebraic analogues of matrices in arbitrary dimensions, have numerous applications to data processing tasks in computer science and computational science more generally.
Many tensors used in diverse application domains are sparse, typically containing more than 90\% zero entries.
Efficient computation with sparse tensors hinges on algorithm that can leverage the sparsity to do less work, but the irregular locations of the nonzero entries poses significant challenges to performance engineers.
Many tensor operations such as tensor-vector multiplications can be sped up substantially by finding an appropriate \emph{blocking} of the tensor, and performing operations in each block using carefully tuned code.
However, selecting an efficient blocking must be selected from among many possibilities is computationally challenging.
Previously, Vuduc et al. \cite{VuducDeYe02, VuducDeYe05}  described the \emph{fill} of a sparse tnesor to be the ratio of the number of entries in blocks with at least one nonzero entry to the number of nonzeros, and showed that it is an effective, efficient heuristic for evaluating the quality of a particular blocking scheme.
However, they gave no theoretical performance bounds for their method, and it is vulnerable to several classes of adversarial examples.

In this paper, we present a sampling-based method for finding $(1 + \eps)$-approximation to the fill of a matrix, with probability at least $1 - \delta$, using $O(\log(1/\delta)/\eps^2)$ samples.
We extend our concentration bounds to a more efficient scheme based on sampling without replacement, using the recent Hoeffding-Serfling inequality \cite{BF2015}.
We then implement our scheme and evaluate it on sparse matrices from the University of Florida collection and compare our scheme to that of Vuduc, as implemented in the Optimized Sparse Kernel Interface (OSKI) library, and a brute-force method for obtaining the ground truth. We find that our algorithm provides faster estimates of the fill at all accuracy levels, providing evidence that is both a theoretic and practical improvement.
\end{abstract}

\input{intro}
\input{notation}  
\input{problem}
\input{related}
\input{algorithm}
\input{analysis}
\input{implementation}
\input{conclusion}

%% Bibliography
\bibliographystyle{abbrv}
%\bibliographystyle{abbrv}
\bibliography{bib-all}

\end{document}


% Provides commands to make subfigures (figures with (a), (b) and (c))
\usepackage{subfigure}

% Typesets URLs sensibly - with tt font, clickable in PDFs, and not breaking across lines
\usepackage{url}

% Makes references hyperlinks in PDF output
\usepackage{hyperref}

% Provides ways to include syntax-highlighted source code
\usepackage{listings}
\lstset{frame=single, basicstyle=\ttfamily}

% Provides good access to colours
\usepackage{color}
\usepackage{xcolor}

% Simple command I defined to allow me to mark TODO items in red
\newcommand{\todo}[1] {\textbf{\textcolor{red}{#1}}}

% Command for vector, matrix, tensor names
\renewcommand{\vec}[1] {\mathbf{#1}}
%\usepackage{esvect}
\newcommand{\Mat}[1] {\mathbf{#1}}
\newcommand{\Ten}[1] {\mathbf{\mathcal{#1}}}
\newcommand{\Pop}[1] {\mathcal{#1}}

% Command for important sets
\newcommand{\F} {\mathbb{F}}
\newcommand{\N} {\mathbb{N}}

% Allows fancy stuff in the page header
\usepackage{fancyhdr}
\pagestyle{fancy}

% Vastly improves the standard formatting of captions
\usepackage[margin=10pt,font=small,labelfont=bf, labelsep=endash]{caption}

% Standard title, author etc.
\title{An Efficient Fill Estimation Algorithm for Sparse Tensors in Blocked Formats}
\author{Peter Ahrens \and Nicholas Schiefer \and Helen Xu}
\date{}
% Put text on the left-hand and right-hand side of the header
%\fancyhead{}
%\lhead{}
%\rhead{}
%\chead{}

\begin{document}

  \maketitle

\input{intro}
\input{notation}  
\input{problem}
\input{related}
\input{algorithm}
\input{analysis}
\input{implementation}
\input{conclusion}

%% Bibliography
\bibliographystyle{abbrv}
%\bibliographystyle{abbrv}
\bibliography{bib-all}

\end{document}
